The Error Estimate of Force Calculation in the Inhomogeneous Molecular Systems

### 2011-11-23 shortRange.inhomo/writeDown/main.tex
add remarks according to Christof. change the grant number of Pingwen

\newcommand{\recheck}[1]{{\color{red} #1}}
\renewcommand{\v}[1]{\textbf{\textit{#1}}}
\renewcommand{\d}[1]{\textsf{#1}}

@@

\title{The Error Estimate of Force Calculation in the Inhomogeneous Molecular Systems}
\author{Han Wang}
\affiliation{LMAM and School of Mathematical
  Sciences, Peking University}
\author{Pingwen Zhang}
% \email{pzhang@pku.edu.cn}
\affiliation{LMAM and School of Mathematical
  Sciences, Peking University}

@@

The widely accepted definition of the force error is the \emph{root
  mean square} (RMS) \emph{error}, which is the square root of the second
moment of the error force $\mathcal E(\v r) =
\sqrt{\langle\vert\Delta\v F(\v r)\vert^2\rangle}$. This is 
calculated as:
\begin{align} \nonumber
  \langle\vert\Delta\v F(\v r)\vert^2\rangle
  =\,&
  \big\langle\sum_{j,k}\v f^c(\v r - \v r_j)\cdot\v f^c(\v r - \v r_k)\big\rangle \\ \nonumber
  =\,&
  \big\langle\sum_j\vert\v f^c(\v r - \v r_j)\vert^2\big\rangle +
  \big\langle\sum_{j\neq k}\v f^c(\v r - \v r_j)\cdot\v f^c(\v r - \v r_k)\big\rangle \\ \nonumber
  =\,&
  \int_{\mathbb R^3}\vert\v f^c(\v r - \v r')\vert^2\rho(\v r')\,\d d\v r'
  + \\\label{eqn:tmp4}
  \,&
  \int_{\mathbb R^3\times\mathbb R^3}\v f^c(\v r - \v r')\cdot\v f^c(\v r - \v r'')\rho(\v r', \v r'')\,\d d\v r'\d d\v r''.
\end{align}
Where  $\rho(\v r', \v r'')$ is the pair density. By using the identity
\begin{align}  \label{eqn:tmp7}
  \rho(\v r', \v r'')
  = \rho(\v r')\rho(\v r'') + C (\v r', \v r''),
\end{align}
where 
\begin{align}
C (\v r', \v r'') = [\,\rho(\v r', \v r'') -  \rho(\v r')\rho(\v r'')\,]
\end{align}
denotes the density correlation between position $\v r'$ and $\v
r''$. Eqn. \eqref{eqn:tmp4} becomes
\begin{align} \nonumber
  \langle\vert\Delta\v F(\v r)\vert^2\rangle
  = &\,
  \int_{\mathbb R^3}\vert\v f^c(\v r - \v r')\vert^2\rho(\v r')\,\d d\v r' \,+ \\\nonumber
  &\,
  \bigg[\int_{\mathbb R^3}\v f^c(\v r - \v r')\rho(\v r')\,\d d\v r'\,\bigg]^2 + \\\nonumber
  &\,
  \int_{\mathbb R^3\times\mathbb R^3}\v f^c(\v r - \v r')\cdot\v f^c(\v r - \v r'')\,C(\v r', \v r'')\,\d d\v r'\d d\v r'' \\\label{eqn:tmp5}
  = &\,
  \mathcal E^2_{\textrm{homo}}(\v r) +
%  \langle\Delta\v F(\v r)\,\rangle^2 +
  \mathcal E^2_{\textrm{inhomo}}(\v r) +
  \mathcal E_{\textrm{correlation}}(\v r).
\end{align}
If the system is homogeneous and the particle-particle correlation is
neglected, then only the first term is left on the right hand side of \eqref{eqn:tmp5}, so it
is called the homogeneity error.  This term originates from the
fluctuation of the error force $\Delta\v F(\v r)$.  The second term is
called the inhomogeneity error, because it stems from the density
inhomogeneity of the system. It is worth pointing out that the
inhomogeneity error is nothing but the magnitude of the mean error force,
namely $\mathcal E_{\textrm{inhomo}}(\v r) = \vert\langle\Delta\v F(\v
r)\,\rangle\vert$.  The last term on right hand side of \eqref{eqn:tmp5} originates from the
correlation between the particles, so it is called the correlation
error.
The correlation error involves a double integral, which
implies that the calculation  requires more compuational resources
than the homogeneity and inhomogeneity errors.
The rest of this paper
assumes the 
correlation error does not dominate in the system, and can be safely
neglected.
It is worth pointing out that this is not always true, for example, in
critical region where the corrlation plays an important role in the system,
the corrlation error should be taken into account.

@@

In an inhomogeneous system, the inhomogeneity error may play a
dominate role in the simulation, (see examples in section~\ref{sec:tmp2}).
Since the mean error force can be calculated
by Eqn.~\eqref{eqn:tmp17} on the fly, a simple way to screen out the
inhomogeneity error is to correct the computed force by the mean error
force, namely
\begin{align}\label{eqn:define-fcorr}
  \v F_{\textrm{corr}}(\v r) = \tilde{\v F}(\v r) + \langle\Delta\v F(\v r)\rangle.
\end{align}
In this paper, the mean error force $\langle\Delta\v F(\v r)\rangle$
is also refered as \emph{correction force} in the context of
long-range force correction. This correction works because it takes
into account the force contribution outside the cut-off radius. The mean of
the corrected error force vanishes:
\begin{align}\label{eqn:mean-fcorr}
  \langle\Delta\v F_{\textrm{corr}}(\v r)\rangle
   =
  \big\langle\,
  \v F(\v r) - \tilde{\v F}(\v r) - \langle\Delta\v F(\v r)\rangle\,
  \big\rangle = 0.
\end{align}
The mean square error is given by
\begin{align} \nonumber
  \langle\vert\Delta\v F_{\textrm{corr}}(\v r)\vert^2\rangle
  & =
  \big\langle\,
  [\,\v F(\v r) - \tilde{\v F}(\v r) - \langle\Delta\v F(\v r)\rangle\,]^2
  \big\rangle \\\nonumber
  & =
  \big\langle\,
  [\,\Delta\v F(\v r) - \langle\Delta\v F(\v r)\rangle\,]^2
  \big\rangle \\\nonumber
  & =
  \langle\vert\Delta\v F(\v r)\vert^2\rangle -
  \langle\Delta\v F(\v r)\,\rangle^2 \\
  & =
  \mathcal E^2_{\textrm{homo}}(\v r) +
  \mathcal E_{\textrm{correlation}}(\v r),
\end{align}
containing no inhomogeneity contribution.  It is worth noticing that the mean
square error of the corrected force is nothing but the variance of the
error force: $\langle\vert\Delta\v
F_{\textrm{corr}}(\v r)\vert^2\rangle = \textrm{Var}[\Delta\v
F_{\textrm{corr}}(\v r)] = \textrm{Var}[\Delta\v F(\v r)]$.

@@

@@

\section*{Acknowledgements}
The authors gratefully thank Luigi Delle Site for for useful comments
on the manuscript. The work of P.Z. is supported by grants from
National Natural Science Foundation of China (20490222).

>>>>

\newcommand{\recheck}[1]{{\color{red} #1}}
\newcommand{\redc}[1]{{\color{red} #1}}
\renewcommand{\v}[1]{\textbf{\textit{#1}}}
\renewcommand{\d}[1]{\textsf{#1}}

@@

\title{The Error Estimate of Force Calculation in the Inhomogeneous Molecular Systems}
\author{Han Wang}
\affiliation{LMAM and School of Mathematical
  Sciences, Peking University}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Christof Sch\"utte}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Pingwen Zhang}
% \email{pzhang@pku.edu.cn}
\affiliation{LMAM and School of Mathematical
  Sciences, Peking University}

@@

The widely accepted definition of the force error is the \emph{root
  mean square} (RMS) \emph{error}, which is the square root of the second
moment of the error force $\mathcal E(\v r) =
\sqrt{\langle\vert\Delta\v F(\v r)\vert^2\rangle}$. This is 
calculated as:
\begin{align} \nonumber
  \langle\vert\Delta\v F(\v r)\vert^2\rangle
  =\,&
  \big\langle\sum_{j,k}\v f^c(\v r - \v r_j)\cdot\v f^c(\v r - \v r_k)\big\rangle \\ \nonumber
  =\,&
  \big\langle\sum_j\vert\v f^c(\v r - \v r_j)\vert^2\big\rangle +
  \big\langle\sum_{j\neq k}\v f^c(\v r - \v r_j)\cdot\v f^c(\v r - \v r_k)\big\rangle \\ \nonumber
  =\,&
  \int_{\mathbb R^3}\vert\v f^c(\v r - \v r')\vert^2\rho(\v r')\,\d d\v r'
  + \\\label{eqn:tmp4}
  \,&
  \int_{\mathbb R^3\times\mathbb R^3}\v f^c(\v r - \v r')\cdot\v f^c(\v r - \v r'')\rho(\v r', \v r'')\,\d d\v r'\d d\v r''.
\end{align}
Where  $\rho(\v r', \v r'')$ is the pair density. By using the identity
\begin{align}  \label{eqn:tmp7}
  \rho(\v r', \v r'')
  = \rho(\v r')\rho(\v r'') + C (\v r', \v r''),
\end{align}
where 
\begin{align}
C (\v r', \v r'') = [\,\rho(\v r', \v r'') -  \rho(\v r')\rho(\v r'')\,]
\end{align}
denotes the density correlation between position $\v r'$ and $\v
r''$. Eqn. \eqref{eqn:tmp4} becomes
\begin{align} \nonumber
  \langle\vert\Delta\v F(\v r)\vert^2\rangle
  = &\,
  \int_{\mathbb R^3}\vert\v f^c(\v r - \v r')\vert^2\rho(\v r')\,\d d\v r' \,+ \\\nonumber
  &\,
  \bigg[\int_{\mathbb R^3}\v f^c(\v r - \v r')\rho(\v r')\,\d d\v r'\,\bigg]^2 + \\\nonumber
  &\,
  \int_{\mathbb R^3\times\mathbb R^3}\v f^c(\v r - \v r')\cdot\v f^c(\v r - \v r'')\,C(\v r', \v r'')\,\d d\v r'\d d\v r'' \\\label{eqn:tmp5}
  = &\,
  \mathcal E^2_{\textrm{homo}}(\v r) +
%  \langle\Delta\v F(\v r)\,\rangle^2 +
  \mathcal E^2_{\textrm{inhomo}}(\v r) +
  \mathcal E_{\textrm{correlation}}(\v r).
\end{align}
If the system is homogeneous and the particle-particle correlation is
neglected, then only the first term is left on the right hand side of \eqref{eqn:tmp5}, so it
is called the homogeneity error.  This term originates from the
fluctuation of the error force $\Delta\v F(\v r)$.  The second term is
called the inhomogeneity error, because it stems from the density
inhomogeneity of the system. It is worth pointing out that the
inhomogeneity error is nothing but the magnitude of the mean error force,
namely $\mathcal E_{\textrm{inhomo}}(\v r) = \vert\langle\Delta\v F(\v
r)\,\rangle\vert$.  The last term on right hand side of \eqref{eqn:tmp5} originates from the
correlation between the particles, so it is called the correlation
error.
The correlation error involves a double integral, which
implies that the calculation  requires more computational resources
than the homogeneity and inhomogeneity errors.
The rest of this paper
assumes the 
correlation error does not dominate in the system, and can be safely
neglected.
It is worth pointing out that this is not always true, for example, in
critical region where the correlation plays an important role in the system,
the correlation error should be taken into account.

@@

In an inhomogeneous system, the inhomogeneity error may play a
dominate role in the simulation, (see examples in section~\ref{sec:tmp2}).
Since the mean error force can be calculated
by Eqn.~\eqref{eqn:tmp17} on the fly, a simple way to screen out the
inhomogeneity error is to correct the computed force by the mean error
force, namely
\begin{align}\label{eqn:define-fcorr}
  \v F_{\textrm{corr}}(\v r) = \tilde{\v F}(\v r) + \langle\Delta\v F(\v r)\rangle.
\end{align}
In this paper, the mean error force $\langle\Delta\v F(\v r)\rangle$
is also referred as \emph{correction force} in the context of
long-range force correction. This correction works because it takes
into account the force contribution outside the cut-off radius. The mean of
the corrected error force vanishes:
\begin{align}\label{eqn:mean-fcorr}
  \langle\Delta\v F_{\textrm{corr}}(\v r)\rangle
   =
  \big\langle\,
  \v F(\v r) - \tilde{\v F}(\v r) - \langle\Delta\v F(\v r)\rangle\,
  \big\rangle = 0.
\end{align}
The mean square error is given by
\begin{align} \nonumber
  \langle\vert\Delta\v F_{\textrm{corr}}(\v r)\vert^2\rangle
  & =
  \big\langle\,
  [\,\v F(\v r) - \tilde{\v F}(\v r) - \langle\Delta\v F(\v r)\rangle\,]^2
  \big\rangle \\\nonumber
  & =
  \big\langle\,
  [\,\Delta\v F(\v r) - \langle\Delta\v F(\v r)\rangle\,]^2
  \big\rangle \\\nonumber
  & =
  \langle\vert\Delta\v F(\v r)\vert^2\rangle -
  \langle\Delta\v F(\v r)\,\rangle^2 \\
  & =
  \mathcal E^2_{\textrm{homo}}(\v r) +
  \mathcal E_{\textrm{correlation}}(\v r),
\end{align}
containing no inhomogeneity contribution.  It is worth noticing that the mean
square error of the corrected force is nothing but the variance of the
error force: $\langle\vert\Delta\v
F_{\textrm{corr}}(\v r)\vert^2\rangle = \textrm{Var}[\Delta\v
F_{\textrm{corr}}(\v r)] = \textrm{Var}[\Delta\v F(\v r)]$.

@@

\redc{ The precise calculation of the correction force needs the
  precise knowledge of the density profile of the system (see
  Eqn. \eqref{eqn:tmp3}). Then the corrected force, in turn, improves
  the accuracy of the density profile. It is possible that this
  feedback loop does not converge, namely the system is driven to a
  totally wrong state by the corresponding incorrect correction
  force. It may happen when the simulation starts very far from the
  equilibrium, or when the fluctuation in the system is so large that
  an accurate calculation of the density profile is impossible.
  % In
  % the latter case, the MD simulation are not trivial anyway, for
  % example, the simulation of critical point.
  % For these two cases, the traditional MD simulation is also not easy
  % to handel.
  In our
  simulations, we do not observe any failure of convergence. Firstly,
  the system is far from the critical point, and we always let the
  simulation starts form a good enough configuration, for example, a
  snapshot of a simulation with the uniform small cut-off and lower
  temperature, or a snapshot from another force correction simulation.
  % or the state of the system changes too fast
  % comparing with the correction force updating fequency
  We also observe that the equilibration of the force-corrected
  system is always as fast as the system simulated by uniform and
  adaptive cut-off methods. That means, at least in our simulation
  cases, the force correction does not lead to more effort of 
  equilibrating the system.
}

@@

\section*{Acknowledgments}
The authors gratefully thank Luigi Delle Site for for useful comments
on the manuscript. The work of P.Z. is supported by grants from
National Natural Science Foundation of China (50930003).
